---
title: "Predicting Customer Churn at a Telecom Company"
subtitle: "A logistic regression analysis"
author: "V. Ruzicka"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: flatly
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 4.5, dpi = 110
)
library(tidyverse)
library(broom)
library(pROC)
library(scales)
theme_set(theme_minimal(base_size = 12))
set.seed(2024)   # reproducibility
```

# Executive summary

Churn is expensive. Winning a new subscriber costs far more than keeping one
you already have, so every customer who leaves for a competitor takes real
margin with them. This analysis uses a public dataset of **7,032 telecom
customers** to build a **logistic regression** model that predicts who is
likely to churn and explains why.

```{r load, include=FALSE}
source("R/load_data.R")
churn <- load_churn()
overall_rate <- mean(churn$churn)
```

The short version: most churn traces back to two things, and the business can
act on both of them. Contract type matters most. A two-year contract cuts the
odds of churn by roughly 75% against month-to-month, and the risk keeps falling
with every extra month a customer stays. The overall churn rate sits at
**`r percent(overall_rate, accuracy = 0.1)`**, imbalanced but not lopsided.

The rest is smaller stuff around the edges. Electronic-check payment, paperless
billing, multiple phone lines, and senior-citizen status each nudge risk up by a
modest, statistically significant amount. Monthly charges and fiber-optic
internet are the interesting case: they *look* dangerous in the raw data, then
wash out once tenure and total charges enter the model, because all three are
carrying the same signal. The interpretation section pulls that apart.

On a held-out test set the model lands at an **AUC of about 0.84**, enough to
rank customers by risk and hand a retention team a list worth working.

# The data

The [Telco Customer Churn dataset](https://www.kaggle.com/datasets/blastchar/telco-customer-churn)
contains one row per customer with their account details, the services they
subscribe to, and whether they churned in the last month. The loading and
cleaning logic lives in `R/load_data.R` (downloaded and cached automatically).

```{r glimpse}
glimpse(churn)
```

Most of the cleaning in `R/load_data.R` is routine factor conversion. The one
judgment call worth flagging is `TotalCharges`: it arrives as text because 11
brand-new customers have a blank value, so coercing it to numeric turns those
into `NA`, and I drop them. At 0.2% of the data, that's a rounding error.
`SeniorCitizen` also gets recoded from 0/1 to a readable factor, and `churn`
becomes a 0/1 column for modeling.

# Exploratory data analysis

Before fitting anything, it pays to see how churn moves with a few key
predictors. The patterns mostly foreshadow what the model finds, with one
instructive exception.

## How many customers churn?

```{r eda-rate}
churn |>
  count(Churn) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(Churn, pct, fill = Churn)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = percent(pct, accuracy = 0.1)), vjust = -0.3) +
  scale_y_continuous(labels = percent) +
  labs(title = "Overall churn rate", x = NULL, y = "Share of customers") +
  guides(fill = "none")
```

About one in four customers churned. That imbalance is mild, but it's enough to
make raw accuracy misleading. A model that just guessed "no churn" for everyone
would already score about `r percent(1 - overall_rate, accuracy = 1)`, so the
evaluation below leans on AUC and the confusion matrix instead.

## Churn by contract type

```{r eda-contract}
churn |>
  group_by(Contract) |>
  summarise(churn_rate = mean(churn), n = n()) |>
  ggplot(aes(reorder(Contract, churn_rate), churn_rate, fill = Contract)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = percent(churn_rate, accuracy = 1)), hjust = -0.15) +
  coord_flip(clip = "off") +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Churn rate by contract type", x = NULL, y = "Churn rate") +
  guides(fill = "none")
```

There's the whole analysis in one chart. **Month-to-month customers churn far
more than anyone on a contract.** Getting customers onto a longer term is the
most obvious lever the business has, and the model will back this up.

## Tenure and monthly charges

```{r eda-tenure}
churn |>
  ggplot(aes(tenure, fill = Churn)) +
  geom_histogram(bins = 30, position = "identity", alpha = 0.6) +
  labs(title = "Churn is concentrated among new customers",
       x = "Tenure (months)", y = "Customers", fill = "Churned?")
```

```{r eda-charges}
churn |>
  ggplot(aes(Churn, MonthlyCharges, fill = Churn)) +
  geom_boxplot(width = 0.5, alpha = 0.7) +
  labs(title = "Churners pay more per month", x = NULL, y = "Monthly charges ($)") +
  guides(fill = "none")
```

Two things stand out. Churn is heavily concentrated in a customer's first few
months, and churners tend to pay more each month. Tenure goes on to be one of
the model's strongest predictors. The monthly-charges signal does not: it mostly
evaporates once the other variables are in the mix. That's the instructive
exception flagged above, and a reminder that a single-variable chart can point
the wrong way.

# Modeling

## Train/test split

We hold out 25% of customers as a test set the model never sees during
training, so our performance estimate reflects how the model would behave on
genuinely new customers.

```{r split}
n         <- nrow(churn)
train_idx <- sample(n, size = floor(0.75 * n))
train     <- churn[train_idx, ]
test      <- churn[-train_idx, ]

c(train = nrow(train), test = nrow(test))
```

## Fitting a logistic regression

We regress churn on every available predictor. Logistic regression gets the
first look for one reason above all: it's **interpretable**. Every coefficient
becomes a plain-English statement about odds, which is what a business audience
actually wants. It happens to be well-behaved and hard to overfit on data this
size too, but interpretability is why it's here.

```{r fit}
model <- glm(churn ~ . - Churn,        # all predictors except the text label
             data = train, family = binomial)
```

```{r fit-summary}
glance(model) |> select(null.deviance, deviance, AIC, df.residual)
```

# Evaluation

## Discrimination: ROC and AUC

AUC (area under the ROC curve) measures how well the model ranks a random
churner above a random non-churner. 0.5 is a coin flip; 1.0 is perfect.

```{r roc}
test_prob <- predict(model, newdata = test, type = "response")
roc_obj   <- roc(test$churn, test_prob, quiet = TRUE)
auc_val   <- as.numeric(auc(roc_obj))

ggroc(roc_obj, color = "#2c7fb8", linewidth = 1) +
  geom_abline(slope = 1, intercept = 1, linetype = "dashed", color = "grey60") +
  labs(title = sprintf("ROC curve (test set AUC = %.3f)", auc_val),
       x = "Specificity", y = "Sensitivity")
```

An AUC of **`r sprintf("%.3f", auc_val)`** indicates strong discrimination. The
model is genuinely useful for ranking customers by risk.

## Classification at a chosen threshold

To turn probabilities into yes/no predictions we pick a threshold. The default
0.5 maximizes accuracy, but for retention we often care more about *catching*
churners (recall), so a business might lower it. Here we report the 0.5
confusion matrix:

```{r confusion}
test_pred <- factor(if_else(test_prob > 0.5, "Yes", "No"), levels = c("No", "Yes"))
cm <- table(Predicted = test_pred, Actual = test$Churn)
cm

accuracy    <- sum(diag(cm)) / sum(cm)
sensitivity <- cm["Yes", "Yes"] / sum(cm[, "Yes"])   # recall on churners
specificity <- cm["No",  "No"]  / sum(cm[, "No"])
tibble(accuracy, sensitivity, specificity) |>
  mutate(across(everything(), ~ percent(.x, accuracy = 0.1)))
```

# Interpreting the drivers of churn

Here's the payoff promised earlier. Exponentiate a coefficient and you get an
**odds ratio**: the factor by which the odds of churn change for a one-unit (or
one-category) move in that predictor, with everything else held fixed. Above 1
means more risk, below 1 means less.

```{r odds-ratios}
or_table <- tidy(model, exponentiate = TRUE, conf.int = TRUE) |>
  filter(term != "(Intercept)", p.value < 0.05) |>
  arrange(desc(estimate)) |>
  select(term, odds_ratio = estimate, conf.low, conf.high, p.value)

or_table |>
  mutate(across(c(odds_ratio, conf.low, conf.high), ~ round(.x, 2)),
         p.value = signif(p.value, 2))
```

```{r odds-plot, fig.height=6}
or_table |>
  mutate(term = reorder(term, odds_ratio)) |>
  ggplot(aes(odds_ratio, term)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey60") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), color = "#2c7fb8") +
  scale_x_log10() +
  labs(title = "Significant churn drivers (odds ratios, log scale)",
       subtitle = "Right of the dashed line = increases churn risk",
       x = "Odds ratio (95% CI)", y = NULL)
```

The dominant drivers line up with intuition: **month-to-month contracts and
short tenure** carry the most risk, while one- and two-year contracts are
strongly protective (a two-year contract cuts the odds of churn by roughly
75%). Electronic-check payment, paperless billing, multiple phone lines, and
senior-citizen status each add smaller but significant risk.

One result deserves a second look: **`MonthlyCharges` and fiber-optic internet
are *not* statistically significant in the model**, even though the EDA showed
churners paying more. These billing variables are highly collinear.
`TotalCharges` is essentially tenure times `MonthlyCharges`, and the
fiber-optic customers are the same people paying the most, so once tenure and
total charges are in the model they absorb that signal. A coefficient measured
alongside other variables answers a different question than a raw EDA chart,
which is why "price doesn't matter" would be the wrong conclusion to draw from a
single p-value. The right next step is to address the collinearity (drop the
redundant billing terms or combine them) before making any pricing decision.

# Recommendations

1. **Convert month-to-month customers to longer contracts.** This is the
   largest, most actionable lever. Targeted incentives (discounts, perks) for
   signing a one- or two-year term should pay for themselves.
2. **Intervene early.** Churn risk is highest in the first months. A structured
   onboarding / early-life retention program targets customers when they are
   most vulnerable.
3. **Reduce payment friction.** Electronic-check payment and paperless billing
   are significant, independent risk factors, so a nudge toward automatic
   payment methods is worth an A/B test. (Fiber-optic customers churn more in the raw data,
   but that is collinear with price; investigate it via the billing variables,
   not as a standalone effect.)
4. **Operationalize the model.** Score the active customer base monthly and
   route the top-risk decile to the retention team. The AUC of
   `r sprintf("%.2f", auc_val)` means this list will be meaningfully better
   than random outreach.

# Limitations & next steps

- This is observational data: the odds ratios describe **associations**, not
  proven causes. A real retention experiment (A/B test) would be needed to
  confirm that, e.g., pushing contracts *causes* lower churn.
- Logistic regression assumes roughly additive effects on the log-odds scale.
  A natural next step is to compare against a tree-based model (random forest /
  gradient boosting) to check whether interactions or non-linearities improve
  predictive performance, at some cost to interpretability.
- Threshold choice should be tied to the actual economics (cost of a retention
  offer vs. value of a saved customer), which would let us optimize expected
  profit rather than a generic metric.

---

*Reproduce this analysis end-to-end by cloning the repo, installing `tidyverse`,
`broom`, `pROC`, `scales`, and `rmarkdown`, and rendering this file with*
`rmarkdown::render("churn-analysis.Rmd")`.
